---
title: "Predicting new COVID-19 cases in the US: A Comparison of a Compatmental model (SIR) and XGBOOST Machine Learning Model"
author: "Basil Okola"
date: today
date-format: "MMMM DD, YYYY"
abstract: |
  **Background**: Figuring out the number of new COVID-19 cases was key in aligning interventions to hotspots. This included ensuring there were enough beds for admissions, key protection gear like face masks, oxygen supply and intensive care unit space. A lot of mechanistic models were build borrowing from the traditional SIR model. Additionally, researchers explored machine learning techniques to predict the course of the infections.
  
  **Methods**: In this report, we compare the performance of a SIR model to an XGBOOST prediction model in predicting new cases in each of the American states, 3 years after the pandemic.
  
  **Results**: The XGBOOST ML approach demonstrated robust goodness-of-fit, effectively capturing the non-linear peak dynamics across varying geographic replicates.
format: 
  pdf:
    keep-tex: true
    number-sections: true
    geometry: margin=1in
    header-includes:
      - \usepackage{sectsty}
      - \usepackage{xcolor}
      - \usepackage{fancyhdr}
      - \definecolor{wdlblue}{HTML}{012D4A}
      - \definecolor{wdlteal}{HTML}{00A7B5}
      - \sectionfont{\color{wdlblue}\uppercase}
      - \subsectionfont{\color{wdlblue}}
      - \usepackage{etoolbox}
      - \makeatletter
      - \patchcmd{\@sect}{\fi#7}{\fi\color{wdlteal}\hrule height 1.5pt \kern 3pt\color{black}#7}{}{}
      - \makeatother
      - \pagestyle{fancy}
      # - \fancyhead[L]{The Lancet Infectious Diseases Format}
      # - \fancyhead[R]{WDL Analytics Report}
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
bibliography: reference.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/dependent/the-lancet-infectious-diseases.csl
---


# Introduction

We chose to use data on COVID-19 cases assembled by a team from [The London School of Health & Tropical Medicine](https://vac-lshtm.shinyapps.io/ncov_tracker/#) who built a global COVID-19 tracker shiny app. I was already working on a personal project that modularized parts of the said app into a golem framework and so it was easier adding a prediction modelling capabilities rather than starting a new project. The [COVID19Dash](https://github.com/Bokola/COVID19Dash) app can be cloned from github and run seamlessly. For  predictions look at the `Prediction Model` tab.

# Methods

## Data

We used `cv_states` data that is part of the datasets in the `COVID19Dash` package. It compiles daily COVID-19 cases in each of the states in the US. It also comes with limited spatial support (longitudes & latitudes). We chose `new_cases` as the outcome variable and generated rolling means for 7, 14, and 30 days as predictors in addition to the frequency of each state in the data to avoid having to encode the categorical variable.

## Models

## Missing data

## Accounting for spatial and tempororal structure


```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
library(xgboost)
library(deSolve)

wdl_blue   <- "#012d4a"
wdl_teal   <- "#00a7b5"
wdl_gray   <- "#58595b"
wdl_light  <- "#f4f7f9"
msf_orange <- "#FF5F00"
msf_red <- "#EB001B"

# dummy data for demonstration
data_sample <- data.frame(
  date = seq(as.Date("2025-01-01"), by = "day", length.out = 100),
  new_cases = cumsum(runif(100, 5, 20)),
  population = 1000000,
  latitude = 1.29,
  longitude = 36.82
)


```


```{r}
# modelling_logic.R
# contains core functions for sir and ml modeling

run_modelling_engine <- function(df) {
  # setup colors as constants for use in plots
  cols <- list(
    blue = "#012d4a", teal = "#00a7b5", gray = "#58595b",
    light = "#f4f7f9", orange = "#FF5F00", red = "#EB001B"
  )
  
  # data prep
  df_clean <- df %>%
    mutate(date = as.Date(date),
           time_index = as.numeric(date - min(date)),
           case_ratio = new_cases / population) %>%
    arrange(date)
  
  # xgboost modelling
  feats <- c("latitude", "longitude", "time_index") 
  x <- as.matrix(df_clean[, feats])
  y <- df_clean$new_cases
  
  model_xgb <- xgboost::xgb.train(
    params = list(objective = "reg:squarederror", learning_rate = 0.1),
    data = xgboost::xgb.DMatrix(x, label = y), 
    nrounds = 25, verbose = 0
  )
  df_clean$preds_xgb <- predict(model_xgb, x)
  
  # sir modelling
  y0 <- c(S = 0.999, I = max(df_clean$case_ratio[1], 1e-5), R = 0)
  fit <- optim(
    par = c(beta = 0.25, gamma = 0.1),
    fn = function(p) {
      out <- try(deSolve::ode(y = y0, times = df_clean$time_index, func = function(t,v,p) {
        list(c(-p[1]*v[1]*v[2], p[1]*v[1]*v[2] - p[2]*v[2], p[2]*v[2]))
      }, parms = p), silent = TRUE)
      if (inherits(out, "try-error")) return(1e12)
      sum((out[,3] - df_clean$case_ratio)^2, na.rm = TRUE)
    },
    method = "L-BFGS-B", lower = c(0.01, 0.01), upper = c(2.0, 1.0)
  )
  
  sir_res <- as.data.frame(deSolve::ode(y = y0, times = df_clean$time_index, 
                                        func = function(t,v,p) list(c(-p[1]*v[1]*v[2], p[1]*v[1]*v[2] - p[2]*v[2], p[2]*v[2])), 
                                        parms = fit$par))
  df_clean$preds_sir <- sir_res[,3] * df_clean$population[1]
  
  return(list(data = df_clean, params = fit$par, colors = cols))
}

# run analysis
results <- run_modelling_engine(data_sample)
df <- results$data
cols <- results$colors
```

Model comparisons

```{r}
ggplot(df, aes(x = date)) +
  geom_point(aes(y = new_cases), color = cols$gray, alpha = 0.4) +
  geom_line(aes(y = preds_xgb, color = "XGBoost (ML)"), size = 1) +
  geom_line(aes(y = preds_sir, color = "SIR (Mechanistic)"), size = 1, linetype = "dashed") +
  scale_color_manual(values = c("XGBoost (ML)" = cols$blue, "SIR (Mechanistic)" = cols$orange)) +
  theme_minimal() +
  labs(title = "Daily Case Projections",
       subtitle = "Comparison of ML and SIR outputs",
       y = "New Cases", x = "Date", color = "Model Type") +
  theme(
    plot.title = element_text(color = cols$blue, face = "bold"),
    legend.position = "bottom"
  )
```

Summary of parameters

```{r}
param_df <- data.frame(
  Parameter = c("Beta (Transmission)", "Gamma (Recovery)", "Basic Reproduction (R0)"),
  Value = c(results$params[1], results$params[2], results$params[1]/results$params[2])
)

param_df %>%
  kbl(booktabs = TRUE, linesep = "", digits = 4, caption = "SIR Model Parameters") %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  row_spec(0, background = cols$blue, color = "white", bold = TRUE) %>%
  column_spec(1, bold = TRUE, color = cols$blue) %>%
  column_spec(2, color = cols$orange, bold = TRUE)
```

Simulations

```{r}
# modelling_logic.R
# contains data simulation and model fitting functions

simulate_cv_data <- function(n_states = 5, days = 120) {
  # simulate data matching cv_states structure
  states <- paste0("state_", 1:n_states)
  
  do.call(rbind, lapply(states, function(s) {
    # generate a random growth curve for each replicate
    t <- 1:days
    peak <- runif(1, 40, 80)
    # mechanistic curve with random noise
    cases <- (peak * exp(-0.05 * (t - 60)^2 / 10)) + rnorm(days, 0, 2)
    cases <- pmax(0, cases) 
    
    data.frame(
      state = s,
      date = seq(as.Date("2025-01-01"), by = "day", length.out = days),
      new_cases = round(cases),
      population = 1000000,
      latitude = runif(1, -1, 1),
      longitude = runif(1, 35, 38)
    )
  }))
}

fit_xgb_assessment <- function(df) {
  # clean and feature engineer
  df_proc <- df %>%
    group_by(state) %>%
    mutate(
      time_index = as.numeric(date - min(date)),
      # create lag features for better ml fit
      lag_1 = lag(new_cases, 1, default = 0),
      roll_7 = zoo::rollmean(lag_1, 7, fill = 0, align = "right")
    ) %>%
    ungroup()
  
  # split for assessment
  train_data <- df_proc %>% filter(time_index < max(time_index) - 14)
  test_data  <- df_proc %>% filter(time_index >= max(time_index) - 14)
  
  feats <- c("latitude", "longitude", "time_index", "lag_1", "roll_7")
  
  # train xgboost
  dtrain <- xgboost::xgb.DMatrix(as.matrix(train_data[, feats]), label = train_data$new_cases)
  model <- xgboost::xgb.train(
    params = list(objective = "reg:squarederror", learning_rate = 0.05, max_depth = 4),
    data = dtrain, nrounds = 50, verbose = 0
  )
  
  # predict on test set
  test_preds <- predict(model, as.matrix(test_data[, feats]))
  
  # calculate metrics
  metrics <- test_data %>%
    mutate(pred = test_preds) %>%
    summarise(
      rmse = sqrt(mean((new_cases - pred)^2)),
      mape = mean(abs((new_cases - pred) / (new_cases + 1))) * 100
    )
  
  return(list(metrics = metrics, full_data = df_proc, model = model, feats = feats))
}
```


```{r}
# simulate 10 replicates
sim_data <- simulate_cv_data(n_states = 10, days = 150)
results <- fit_xgb_assessment(sim_data)

results$metrics %>%
  kbl(booktabs = TRUE, caption = "Global Assessment Metrics") %>%
  kable_styling(latex_options = "hold_position") %>%
  row_spec(0, background = wdl_blue, color = "white")

sample_state <- results$full_data %>% 
  filter(state == "state_1") %>%
  mutate(pred = predict(results$model, as.matrix(.[, results$feats])))

ggplot(sample_state, aes(x = date)) +
  geom_point(aes(y = new_cases), color = wdl_gray, alpha = 0.5) +
  geom_line(aes(y = pred), color = msf_orange, size = 1) +
  theme_minimal() +
  labs(title = "XGBoost Fit vs Simulated Data",
       subtitle = "State 1 - Orange line represents model prediction",
       x = "Date", y = "New Cases")
```

